---
title: "Hagfish length, weight, diameter relationships"
author: "Ben Williams, Jane Sullivan"
date: "January 24, 2019"
output:
  bookdown::pdf_document2: 
  fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
```

```{r}
source('code/helper.r')
library(knitr)
options(scipen = 999)
hag <- read_csv('data/hag_lwd.csv')
```

## Questions {-}

 - calculate the diameter of a hagfish based upon length and weight,
 - calculate the hole size that a hagfish can squeeze through.
 

## Exploratory data analysis {-}

First a couple of basic examinations of the data.

```{r lw, fig.align='center', fig.cap='Hagfish length-weight measurements.'}
hag %>% 
  ggplot(aes(length, weight)) + geom_point() +
  expand_limits(x = 0, y = 0) +
  xlab('\nLength (cm)') + 
  ylab('Weight (g)\n') +
  stat_smooth()
```

```{r ld, fig.align='center', fig.cap='Hagfish length-diameter measurements.'}
hag %>% 
  ggplot(aes(length, diameter)) + geom_point() +
  expand_limits(x = 0, y = 0) +
  xlab('\nLength (cm)') + 
  ylab('Diameter (cm)\n') +
  stat_smooth()
```

```{r wd, fig.align='center', fig.cap='Hagfish weight-diameter measurements.'}
hag %>% 
  ggplot(aes(weight, diameter)) + geom_point() +
  expand_limits(x = 0, y = 0) +
  xlab('\nWeight (g)') + 
  ylab('Diameter (cm)\n') +
  stat_smooth()

```


## Models {-}

A suite of models were examined to explore diameter ~ length/weight relationships.
Generalized additive models and generalized linear models with and without interaction and/or log links were examined.

```{r, echo = T}
gam_l <- gam(diameter ~ s(length, k=4), data = hag, gamma = 1.4, method='ML')
gam_w <- gam(diameter ~ s(weight, k=4), data = hag, gamma = 1.4, method='ML')
gam_lw <- gam(diameter ~ s(length, k=4) + s(weight, k=4), data = hag, gamma = 1.4, method='ML')
gam_llw <- gam(diameter ~ s(length, k=4) + s(weight, k=4), data = hag, gamma = 1.4, method='ML', family = gaussian(link = 'log'))


glm_l <- glm(diameter ~ length, data = hag)
glm_w <- glm(diameter ~ weight, data = hag)
glm_lw <- glm(diameter ~ length * weight, data = hag)
glm_llw <- glm(diameter ~ length * weight, data = hag, family = gaussian(link = 'log'))

null_lm <- lm(diameter ~ length, data = hag)
```

AIC was used to test for the top model.

```{r}
AIC(gam_l, gam_w, gam_lw, gam_llw, glm_l, glm_w, glm_lw, glm_llw, null_lm)
```


The top model is the glm with an interaction term between length and weight (glm_lw). 
Summary of model results.

```{r}
summary(glm_lw)
```

Plots of model results

```{r}
plot(glm_lw)
```

# Comparisons {-}

Compare best fit model to the Harada estimate (a linear model) and a linear model using ADFG data. This was explored by examining model residuals by length and weight.

```{r, fig.align='center', fig.cap='Hagfish model residuals by weight.'}
hag %>% 
  mutate(adfg_best_model = fitted(glm_lw),
         harada_linear = (length * 0.155 + 0.049) / 3.1416,
         # linear model using our data
         adfg_linear = fitted(null_lm),
         resid_best = diameter - adfg_best_model,
         resid_harada_linear = diameter - harada_linear,
         resid_linear = diameter - adfg_linear) %>% 
  gather(variable, resid, -length, -weight, -diameter, -adfg_best_model, -harada_linear, - adfg_linear) %>% 
  ggplot(aes(weight, resid, color = variable, fill = variable )) + 
  geom_point() +
  geom_hline(yintercept=0, lty = 3) +
  stat_smooth(alpha = .2)
```


```{r, fig.align='center', fig.cap='Hagfish model residuals by length'}
hag %>% 
  mutate(adfg_best_model = fitted(glm_lw),
         harada_linear = (length * 0.155 + 0.049) / 3.1416,
         # linear model using our data
         adfg_linear = fitted(null_lm),
         resid_best = diameter - adfg_best_model,
         resid_harada_linear = diameter - harada_linear,
         resid_linear = diameter - adfg_linear) %>% 
  gather(variable, resid, -length, -weight, -diameter, -adfg_best_model, -harada_linear, - adfg_linear) %>% 
  ggplot(aes(length, resid, color = variable, fill = variable )) + 
  geom_point() +
  geom_hline(yintercept=0, lty = 3) +
  stat_smooth(alpha = .2)
```

These show that the Harada parameter estimates do a poor job at lengths > 50, and are strongly biased when compared via weight. The best fit model does a good job of describing the output.



```{r, fig.align='center', fig.cap="Observed and predicted hagfish diameters"}
hag %>% 
  mutate(pred = fitted(glm_lw)) %>% 
  ggplot(aes(length, diameter)) + 
  geom_point() +
  geom_point(aes(length, pred), color = 2, shape = 1)
```

The coefficent estimates can now be used as predictors for future samples

```{r, echo = T}
length = seq(20, 60, by = 5)
length = 20
weight = 20
weight = seq(50, 450, by = 50)

coef(glm_lw)[[1]] + coef(glm_lw)[[2]] * length + coef(glm_lw)[[3]] * weight + coef(glm_lw)[[4]] * length * weight

summary(glm_lw)
```

In short it appears that you should be able to use length/weight to get a reasonable estimate of diameter. 
This of course assumes that the sampling was distributed both in space and time - if not you will want to do some additional sampling to beef up this analysis.

# Escape size {-}

To determine escape size I worked up a function that predicts the $L_{50}$ based upon 3 inputs:

-   shrink = the amount a hagfish is estimated to reduce in size (e.g,. 0.15 is assuming a shrinkage of 15%) to fit through an opening
-   escape_size = escape hole diameter (cm)
-   prob = the probability e.g., (0.50 for 50%)

```{r, echo = T}

length_caught <- function(
  shrink, escape_size, prob) {
  
  hag %>%
    mutate(dia = diameter / (1 + shrink),
           catch = factor(ifelse(dia > escape_size, 1, 0))) -> df 
  
    model = glm(catch ~ length, family = 'binomial', data = df)

    t = round((log(prob / (1 - prob)) - coef(model)[1]) / coef(model)[2])
  
    df %>% 
      mutate(pred = predict(model, ., type ='response')) %>% 
      ggplot(aes(length, pred)) + geom_point() +
      ggtitle(paste0('L', prob*100, ' = ', t, ' cm'))
}

length_caught(0.15, 1.9, 0.50)
```

The length that 50% of the hagfish will be retained is reported at the top of the figure. 
The values can be adjusted in case you think that a hagfish can reduce in size by 20% etc.

## Alternative method to get escape size {-}

You could also use $L_{50}$ (or some other reference point) from a maturity curve to determine an optimal escape hole size. Similar to above, this method is based on the estimated shrink, and the probability that you want to base the output on (defaults to 50%). In addition, you put in the $L_{50}$ that is estimated externally from a maturity curve.

```{r, echo = T}

myL50 <- 45 # user input! waiting for the estimated L50 from from Andrew.

f_len_caught <- function(escape_size = 1.9, 
                         shrink = 0.15,
                         prob = 0.5,  
                         target_length = myL50) {
  hag %>% mutate(dia = diameter / (1 + shrink),
                 catch = factor(ifelse(dia > escape_size, 1, 0))) -> df 
  model <- glm(catch ~ length, family = 'binomial', data = df)
  len <- round((log(prob / (1 - prob)) - coef(model)[1]) / coef(model)[2])
  out <- len - target_length # ultimately we want the difference between predicted len and target length to be 0
  return(out)
}

# Finds the escape ring size between the interval provided that gives us a 50% probability of releasing our 50% probability of being mature fish
fit <- uniroot(f = f_len_caught, interval = c(0.9,2.5))
optim_escape <- round(fit$root,1)

```

For example, given that the length at 50% maturity ($L_{50}$) is `r round(myL50)` cm, the optimal escape hole size that results in a 50% probability of retention for $L_{50}$ is `r optim_escape` cm.

## Length predictions from the Harada model for comparison {-}

As requested by Aaron, this function uses the paramter estimates for the logistic curve and linear regression from Harada et al. 2007 to output the length at which some proportion of fish (e.g. 50%) are retained given an escape hole size. The output is the length (cm) at which (prob, %) fish are retained by the gear given an escape hole size from the Harada model. Recall that the Harada model is based on relative girth, so length must be derived from the length/girth regression.

```{r harada, echo = T}

predict_harada <- function(
  # Probability from 0 to 1
  prob = 0.75,
  # Escape hole diameter (cm)
  escape_size = 1.9,
  # Logistic parameter estimates from Table 2 in Harada et al 2007
  logistic_alpha = -13.34,
  logistic_beta = 10.99,
  #length/girth linear regression estimates from Harada. Could use parameter
  #estimates from our best-fitting model instead.
  linreg_b0 = 4.213, # y-intercept
  linreg_b1 = 5.487 # slope
) {
# Relative girth (R) predicted by the model
R <- (log(prob/(1-prob))-logistic_alpha)/logistic_beta
# Girth (G) predicted by the model (adjusted by escape hole perimeter, which is diameter * pi)
G <- R * escape_size * pi
# Length predicted by the model using the length/girth linear regression estimates
L <- linreg_b0 + linreg_b1 * G
return(round(L,2))
}
```


For example, we can use this function to get the predicted lengths given some probability of retention and escape hole size:

```{r test_harada, echo = T}
prob_vec <- c(0.1, 0.25, 0.5, 0.75, 0.9)
escape_vec <- c(1.6, 1.9)

out <- matrix(nrow = length(prob_vec), ncol = length(escape_vec))

for(i in 1:length(prob_vec)){
  for(j in 1:length(escape_vec)){
out[i,j] <- predict_harada(prob = prob_vec[i], escape_size = escape_vec[j])
  }
}

out <- as.data.frame(out, row.names = FALSE)
names(out) <- paste0(escape_vec, " cm")
out %>% 
  mutate(Probability = prob_vec) %>% 
  select(Probability, `1.6 cm`, `1.9 cm`) -> out
```

`r kable(out)`
