---
title: "Hagfish length, weight, diameter relationships"
author: "Ben Williams"
date: "January 17, 2019"
output:
  bookdown::html_document2: 
  fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
```

```{r}
source('code/helper.r')
options(scipen = 999)
hag <- read_csv('data/hag_lwd.csv')
```

## Questions {-}

 - calculate the diameter of a hagfish based upon length and weight,
 - calculate the hole size that a hagfish can squeeze through.
 

## Exploratory data analysis {-}

First a couple of basic examinations of the data.

```{r lw, fig.align='center', fig.cap='Hagfish length-weight measurements.'}
hag %>% 
  ggplot(aes(length, weight)) + geom_point() +
  expand_limits(x = 0, y = 0) +
  xlab('\nLength (cm)') + 
  ylab('Weight (g)\n') +
  stat_smooth()
```

```{r ld, fig.align='center', fig.cap='Hagfish length-diameter measurements.'}
hag %>% 
  ggplot(aes(length, diameter)) + geom_point() +
  expand_limits(x = 0, y = 0) +
  xlab('\nLength (cm)') + 
  ylab('Diameter (cm)\n') +
  stat_smooth()
```

```{r wd, fig.align='center', fig.cap='Hagfish weight-diameter measurements.'}
hag %>% 
  ggplot(aes(weight, diameter)) + geom_point() +
  expand_limits(x = 0, y = 0) +
  xlab('\nWeight (g)') + 
  ylab('Diameter (cm)\n') +
  stat_smooth()

```


## Models {-}

A suite of models were examined to explore diameter ~ length/weight relationships.
Generalized additive models and generalized linear models with and without interaction and/or log links were examined.

```{r, echo = T}
gam_l <- gam(diameter ~ s(length, k=4), data = hag, gamma = 1.4, method='ML')
gam_w <- gam(diameter ~ s(weight, k=4), data = hag, gamma = 1.4, method='ML')
gam_lw <- gam(diameter ~ s(length, k=4) + s(weight, k=4), data = hag, gamma = 1.4, method='ML')
gam_llw <- gam(diameter ~ s(length, k=4) + s(weight, k=4), data = hag, gamma = 1.4, method='ML', family = gaussian(link = 'log'))


glm_l <- glm(diameter ~ length, data = hag)
glm_w <- glm(diameter ~ weight, data = hag)
glm_lw <- glm(diameter ~ length * weight, data = hag)
glm_llw <- glm(diameter ~ length * weight, data = hag, family = gaussian(link = 'log'))
```

AIC was used to test for the top model.

```{r}
AIC(gam_l, gam_w, gam_lw, gam_llw, glm_l, glm_w, glm_lw, glm_llw)
```


The top model is the glm with an interaction term between length and weight (glm_lw). 
Summary of model results.

```{r}
summary(glm_lw)
```

Plots of model results

```{r}
plot(glm_lw)
```

# Comparisons {-}

Compare best fit model to the Harada estimate. 
This was explored by examining model residuals by length and weight.

```{r, fig.align='center', fig.cap='Hagfish model residuals by weight.'}
hag %>% 
  mutate(pred = fitted(glm_lw),
         harada = (length * 0.155 + 0.049) / 3.1416,
         resid = diameter - pred,
         resid_ha = diameter - harada) %>% 
  gather(variable, resid, -length, -weight, -diameter, -pred, -harada) %>% 
  ggplot(aes(weight, resid, color = variable, fill = variable )) + 
  geom_point() +
  geom_hline(yintercept=0, lty = 3) +
  stat_smooth(alpha = .2)
```


```{r, fig.align='center', fig.cap='Hagfish model residuals by length'}
hag %>% 
  mutate(pred = fitted(glm_lw),
         harada = (length * 0.155 + 0.049) / 3.1416,
         resid = diameter - pred,
         resid_ha = diameter - harada) %>% 
  gather(variable, resid, -length, -weight, -diameter, -pred, -harada) %>% 
  ggplot(aes(length, resid, color = variable, fill = variable )) + 
  geom_point() +
  geom_hline(yintercept=0, lty = 3) +
  stat_smooth(alpha = .2)
```

These show that the Harada parameter estimates do a poor job at ~ length 50, and are strongly biased when compared via weight. The best fit model does a good job of describing the output.



```{r, fig.align='center', fig.cap="Observed and predicted hagfish diameters"}
hag %>% 
  mutate(pred = fitted(glm_lw)) %>% 
  ggplot(aes(length, diameter)) + geom_point() +
  geom_point(aes(length, pred), color = 2)
```

The coefficent estimates can now be used as predictors for future samples

```{r}
length = seq(20, 60, by = 5)
weight = seq(50, 450, by = 50)

coef(glm_lw)[[1]] + coef(glm_lw)[[2]] * length + coef(glm_lw)[[3]] * weight + coef(glm_lw)[[4]] * length * weight
```

In short it appears that you should be able to use length/weight to get a reasonable estimate of diameter. 
This of course assumes that the sampling was distributed both in space and time - if not you will want to do some additional sampling to beef up this analysis.

# Escape size {-}

To determine escape size I worked up a function that predicts the L50 based upon 3 inputs:
 - shrink = the amount a hagfish is estimated to reduce in size (e.g,. 0.15) to fit through an opening
 - target_size = the target diameter size (cm)
 - prob = the probability e.g., (0.50 for 50%)

```{r, echo = T}

length_caught <- function(shrink, target_size, prob) {
  
  hag %>%
    mutate(dia = diameter / (1 + shrink),
         catch = factor(ifelse(dia > target_size, 1, 0))) -> df 
  
    model = glm(catch ~ length, family = 'binomial', data = df)

    t = round((log(prob / (1 - prob)) - coef(model)[1]) / coef(model)[2])
  
    df %>% 
      mutate(pred = predict(model, ., type ='response')) %>% 
      ggplot(aes(length, pred)) + geom_point() +
      ggtitle(paste0('L50 = ', t, 'cm'))
}


length_caught(0.15,  2.0, 0.5)
```

The length that 50% of the hagfish will be retained is reported at the top of the figure. 
The values can be adjusted in case you think that a hagfish can reduce in size by 20% etc.



